Recurrent event analysis in the presence of real-time high frequency data via random subsampling

### 2020-05-21 write-up/fda-recurrent.tex
Update fda-recurrent.tex

\def\W{{\bf W}}
\def\Bka{{\it Biometrika}}
\def\AIC{\textsc{aic}}
\def\T{{ \mathrm{\scriptscriptstyle T} }}
\def\v{{\varepsilon}}
\def\partitionsn{\mathop{\mathcal{P}_{[n]}}\nolimits}
\def\partitionsN{\mathop{\mathcal{P}_{\infty}}\nolimits}
\def\Dmn{\mathop{{D}_{m,n}}\nolimits}
\def\symmetricn{\mathop{\mathcal{S}_{n}}\nolimits}
\def\PE{\mathop{\rm Pitman\mbox{-}Ewens}\nolimits}
\def\per{\mathop{\rm per}\nolimits}
\def\U{\mathop{\mathcal{U}_{}}\nolimits}
\def\Nb{\mathop{\mathbb{N}_{}}\nolimits}
\def\Nbb{\mathop{\mathbf{N}_{}}\nolimits}
\def\nbb{\mathop{\mathbf{n}_{}}\nolimits}
\def\Xbb{\mathop{\mathbf{X}_{}}\nolimits}
\def\fin{\mathop{\text{fin}}\nolimits}
\def\pr{\mathop{\text{pr}}\nolimits}
\def\E{\mathcal{E}}
\def\G{\mathcal{G}}
\def\deg{\text{deg}_{}}
\def\equalinlaw{=_{\mathcal{D}}}
\def\Fk{\mathop{\mathcal{F}_k^{\downarrow}}\nolimits}
\def\F{\mathop{\mathcal{F}^{\downarrow}}\nolimits}
\def\size{\mathop{\text{size}}\nolimits}
\def\Ycong{\mathop{\EY}\nolimits}
\def\EIsig{\mathop{\mathcal{E}_{\mathcal{I}}^{\sigma}}\nolimits}
\def\EY{\mathop{\mathcal{E}_{Y}}\nolimits}
\def\finp{\mathop{\fin(\mathcal{P})}\nolimits}
\def\En{\mathop{\mathfrak{E}_{[n]}}\nolimits}
\def\EN{\mathop{\mathfrak{E}_{\Nb}}\nolimits}
\def\logit{\mathop{\text{logit}_{}}\nolimits}
\def\EI{\mathop{\mathcal{E}_{\mathcal{I}}}\nolimits}
\def\ES{\mathop{\mathfrak{E}_S}\nolimits}
\def\Ifk{\mathop{{I}_{}}\nolimits}
%\def\mathcal{I}{\mathop{\mathcal{I}_{}}\nolimits}
\def\Ycong{\mathop{\EY}\nolimits}
\def\EIsig{\mathop{\mathcal{E}_{\mathcal{I}}^{\sigma}}\nolimits}
\def\EY{\mathop{\mathcal{E}_{Y}}\nolimits}
\def\finp{\mathop{\fin(\mathcal{P})}\nolimits}
\def\En{\mathop{\mathfrak{E}_{[n]}}\nolimits}
\def\EN{\mathop{\mathfrak{E}_{\Nb}}\nolimits}
\def\logit{\mathop{\text{logit}_{}}\nolimits}
\def\EI{\mathop{\mathcal{E}_{\mathcal{I}}}\nolimits}
\def\ES{\mathop{\mathfrak{E}_S}\nolimits}

@@

In standard logistic mixed-effects model, with $b_i \sim N(0, \Psi)$ and $b_i \in \mathbb{R}^q$, then the likelihood is given by
$$
\begin{aligned}
& \prod_{i=1}^n \int \prod_{j=1}^{k_i} \left( \frac{\exp(W_{ij} \theta + Z_{ij} b_i + \log \pi_{ij})}{1 + \exp(W_{ij} \beta + Z_{ij} b_i + \log \pi_{ij})} \right)^{y_{ij}} \left( \frac{1}{1 + \exp(W_{ij} \beta + Z_{ij} b_i + \log \pi_{ij})} \right)^{1-y_{ij}} \phi (b_i; \sigma_b^2) db_i \\
\propto & \prod_{i=1}^n \int \exp \left[ \sum_{j=1}^{k_i} \left( y_{ij} \left( W_{ij} \theta + Z_{ij} b_i + \log \pi_{ij} \right) - \log \left( 1 + \exp \left( W_{ij} \theta + Z_{ij} b_i + \log \pi_{ij} \right) \right) \right) - b_i^\top \Psi^{-1} b_i /2 \right] db_i \\
\propto& \prod_{i=1}^n \int \exp \left[ g(\theta, b_i, \pi_{ij}, \Psi) \right] db_i,
\end{aligned}
$$
where the second and third line ignore the constant $(2\pi)^{q/2} |\Psi|^{-1/2}$. Differentiating the function $g(\theta, b_i, \pi_{ij}, \Psi)$ with respect to $b_i$ yields
$$
\begin{aligned}
\frac{\partial g(\theta, b_i, \pi_{ij}, \Psi)}{\partial b_i} &= \sum_{j=1}^{k_i} \left( y_{ij} Z_{ij} - \frac{Z_{ij}}{1+\exp(-W_{ij}^\top \theta - Z_{ij}^\top b_i - \log \pi_{ij}}\right) - \Psi^{-1} b_i \\
\frac{\partial g(\theta, b_i, \pi_{ij}, \Psi)}{\partial b_i^2} &= - \sum_{j=1}^{k_i} \frac{Z_{ij}Z_{ij}^\top}{(W_{ij}^\top \theta + Z_{ij}^\top b_i + \log \pi_{ij})^2} - \Psi^{-1}
\end{aligned}
$$
Redefining terms from the prior section, we can express the above as
$$
\begin{aligned}
\frac{\partial g(\theta, b_i, \pi_{ij}, \Psi)}{\partial b_i} &= Z_i E_i (\theta, b_i) - \Psi^{-1} b_i \\
\frac{\partial g(\theta, b_i, \pi_{ij}, \Psi)}{\partial b_i^2} &= - \left[ Z_i V_i (\theta, b_i) Z_i + \Psi^{-1} \right]
\end{aligned}
$$

For fixed $\theta$, the function $g$ is stictly concave function of $b_i$ and therefore we can estimate the unique maximum $\hat b_i$ via Newton-Rhapson method as follows:
$$
\hat b_i^{(k+1)} = \hat b_i^{(k)} + \left( \sum_{j=1}^{k_i} \frac{Z_{ij}Z_{ij}^\top}{(W_{ij}^\top \theta + Z_{ij}^\top \hat b_i^{(k)} + \log \pi_{ij})^2} + \Psi^{-1} \right)^{-1} \left( \sum_{j=1}^{k_i} \left( y_{ij} Z_{ij} - \frac{Z_{ij}}{W_{ij}^\top \theta + Z_{ij}^\top \hat b_i^{(k)} + \log \pi_{ij}}\right) - \Psi^{-1} \hat b_i^{(k)} \right)
$$
Fixing $\hat b_i$, we then need to estimate $\Psi$ and $\theta$. We employ adaptive gaussian quadrature (cites) as it has been shown that for infrequent events the Laplace approximation tends to underestimate effects (cites).
To this end, we

We write this as $\prod_{i=1}^n l_{ij} (\theta, b_i, \pi_{ij}) $

>>>>

\def\k{{\bf k}}
\def\z{{\bf z}}
\def\W{{\bf W}}
\def\Bka{{\it Biometrika}}
\def\AIC{\textsc{aic}}
\def\T{{ \mathrm{\scriptscriptstyle T} }}
\def\v{{\varepsilon}}
\def\partitionsn{\mathop{\mathcal{P}_{[n]}}\nolimits}
\def\partitionsN{\mathop{\mathcal{P}_{\infty}}\nolimits}
\def\Dmn{\mathop{{D}_{m,n}}\nolimits}
\def\symmetricn{\mathop{\mathcal{S}_{n}}\nolimits}
\def\PE{\mathop{\rm Pitman\mbox{-}Ewens}\nolimits}
\def\per{\mathop{\rm per}\nolimits}
\def\U{\mathop{\mathcal{U}_{}}\nolimits}
\def\Nb{\mathop{\mathbb{N}_{}}\nolimits}
\def\Nbb{\mathop{\mathbf{N}_{}}\nolimits}
\def\nbb{\mathop{\mathbf{n}_{}}\nolimits}
\def\Xbb{\mathop{\mathbf{X}_{}}\nolimits}
\def\fin{\mathop{\text{fin}}\nolimits}
\def\pr{\mathop{\text{pr}}\nolimits}
\def\E{\mathcal{E}}
\def\G{\mathcal{G}}
\def\deg{\text{deg}_{}}
\def\equalinlaw{=_{\mathcal{D}}}
\def\Fk{\mathop{\mathcal{F}_k^{\downarrow}}\nolimits}
\def\F{\mathop{\mathcal{F}^{\downarrow}}\nolimits}
\def\size{\mathop{\text{size}}\nolimits}
\def\Ycong{\mathop{\EY}\nolimits}
\def\EIsig{\mathop{\mathcal{E}_{\mathcal{I}}^{\sigma}}\nolimits}
\def\EY{\mathop{\mathcal{E}_{Y}}\nolimits}
\def\finp{\mathop{\fin(\mathcal{P})}\nolimits}
\def\En{\mathop{\mathfrak{E}_{[n]}}\nolimits}
\def\EN{\mathop{\mathfrak{E}_{\Nb}}\nolimits}
\def\logit{\mathop{\text{logit}_{}}\nolimits}
\def\EI{\mathop{\mathcal{E}_{\mathcal{I}}}\nolimits}
\def\ES{\mathop{\mathfrak{E}_S}\nolimits}
\def\Ifk{\mathop{{I}_{}}\nolimits}
%\def\mathcal{I}{\mathop{\mathcal{I}_{}}\nolimits}
\def\Ycong{\mathop{\EY}\nolimits}
\def\EIsig{\mathop{\mathcal{E}_{\mathcal{I}}^{\sigma}}\nolimits}
\def\EY{\mathop{\mathcal{E}_{Y}}\nolimits}
\def\finp{\mathop{\fin(\mathcal{P})}\nolimits}
\def\En{\mathop{\mathfrak{E}_{[n]}}\nolimits}
\def\EN{\mathop{\mathfrak{E}_{\Nb}}\nolimits}
\def\logit{\mathop{\text{logit}_{}}\nolimits}
\def\EI{\mathop{\mathcal{E}_{\mathcal{I}}}\nolimits}
\def\ES{\mathop{\mathfrak{E}_S}\nolimits}

@@

In standard logistic mixed-effects model, with $b_i \sim N(0, \Psi)$ and $b_i \in \mathbb{R}^q$, then the likelihood is given by
$$
\begin{aligned}
& \prod_{i=1}^n \int \prod_{j=1}^{k_i} \left( \frac{\exp(W_{ij} \theta + Z_{ij} b_i + \log \pi_{ij})}{1 + \exp(W_{ij} \beta + Z_{ij} b_i + \log \pi_{ij})} \right)^{y_{ij}} \left( \frac{1}{1 + \exp(W_{ij} \beta + Z_{ij} b_i + \log \pi_{ij})} \right)^{1-y_{ij}} \phi (b_i; \sigma_b^2) db_i \\
\propto & \prod_{i=1}^n \int \exp \left[ \sum_{j=1}^{k_i} \left( y_{ij} \left( W_{ij} \theta + Z_{ij} b_i + \log \pi_{ij} \right) - \log \left( 1 + \exp \left( W_{ij} \theta + Z_{ij} b_i + \log \pi_{ij} \right) \right) \right) - b_i^\top \Psi^{-1} b_i /2 \right] db_i \\
\propto& \prod_{i=1}^n \int \exp \left[ g(\theta, b_i, \pi_{ij}, \Psi) \right] db_i,
\end{aligned}
$$
where the second and third line ignore the constant $(2\pi)^{q/2} |\Psi|^{-1/2}$. Differentiating the function $g(\theta, b_i, \pi_{ij}, \Psi)$ with respect to $b_i$, the connection to the prior section can be used.  Re-defining terms appropriately allows us to express the derivatives as
$$
\begin{aligned}
\frac{\partial g(\theta, b_i, \pi_{i}, \Psi)}{\partial b_i} &= Z_i E_i (\theta, b_i) - \Psi^{-1} b_i \\
\frac{\partial g(\theta, b_i, \pi_{i}, \Psi)}{\partial b_i^2} &= - \left[ Z_i V_i (\theta, b_i) Z_i + \Psi^{-1} \right]
\end{aligned}
$$
For fixed $\theta$, the function $g$ is stictly concave function of $b_i$ and therefore we can estimate the unique maximum $\hat b_i$ via Newton-Rhapson method as follows:
$$
\hat b_i^{(k+1)} = \hat b_i^{(k)} + \left[ Z_i V_i (\theta, \hat b_i^{(k)}) Z_i + \Psi^{-1} \right]^{-1}  \left( Z_i E_i (\theta, \hat b_i^{(k)}) - \Psi^{-1} \hat b_i^{(k)} \right)
$$

Fixing $\hat b_i$, we then need to estimate $\Psi$ and $\theta$. Here we employ adaptive gaussian quadrature (cites) as it has been shown that for infrequent events the Laplace approximation tends to underestimate effects (cites).  Let $R_i^\top R_i = Z_i^\top V_i (\theta, b_i) Z_i + \Psi^{-1}$, $z_j$ and $w_j$ for $j=1,\ldots, N_{GQ}$ be the weights for the one-dimensional Gaussian quadrature rule.  Let $\z_\k = (z_{k_1},\ldots, z_{k_q})$, $\tilde b_{i \k} = \hat b_i + R_i^{-1} \z_\k$, and $W_{\k} = \exp ( \| \z_\k \|^2 ) \prod_{l=1}^{q} w_{k_l}$.  Then the AGQ rule is given by
$$
\begin{aligned}
&\int \exp \left[ g(\theta, b_i, \pi_{i}, \Psi) \right] db_i \\
=& \int | R_i |^{-1} \exp \left[ g(\theta, \hat b_i + R_i^{-1} \z, \pi_{i}, \Psi) + \| \z \|^2 / 2 \right] \exp ( - \| \z \|^2 / 2 ) d\z \\
=& (2 \pi)^{q/2} | R_i |^{-1} \sum_{j_1 =1}^{N_{GQ}} \cdots \sum_{j_q =1}^{N_{GQ}} \exp \left[ g(\theta, \tilde b_{i \k}, \pi_{i}, \Psi) \right] W_{\k}
\end{aligned}
$$
The score equations
$$
\sum_{j_1 =1}^{N_{GQ}} \cdots \sum_{j_q =1}^{N_{GQ}}
\left( \sum_{j_1 =1}^{N_{GQ}} \cdots \sum_{j_q =1}^{N_{GQ}} \exp \left[ g(\theta, \tilde b_{i \k}, \pi_{i}, \Psi) \right] W_{\k} \right)^{-1}
\left( \exp \left[ g(\theta, \tilde b_{i \k}, \pi_{i}, \Psi) \right] W_{\k} \right) \frac{\partial}{\partial \theta} g(\theta, \tilde b_{i \k}, \pi_i, \Psi)
$$
Define $\omega(\theta, \tilde b_{i \k}, \pi_i, \Psi)$ to be the weights.  Then we can express score and Hessian as:
$$
\begin{aligned}
&\sum_{\k} \omega(\theta, \tilde b_{i \k}, \pi_i, \Psi) \cdot \frac{\partial}{\partial \theta} g(\theta, \tilde b_{i \k}, \pi_i, \Psi) \\
&\sum_{\k}
\omega(\theta, \tilde b_{i \k}, \pi_i, \Psi) \cdot \frac{\partial^2}{\partial \theta \partial \theta^\top} g(\theta, \tilde b_{i \k}, \pi_i, \Psi) +
\sum_{\k} \omega(\theta, \tilde b_{i \k}, \pi_i, \Psi) \cdot \left [\frac{\partial}{\partial \theta} g(\theta, \tilde b_{i \k}, \pi_i, \Psi) \right] \left [\frac{\partial}{\partial \theta} g(\theta, \tilde b_{i \k}, \pi_i, \Psi) \right]^\top \\
&- \left[ \sum_{\k} \omega(\theta, \tilde b_{i \k}, \pi_i, \Psi) \frac{\partial}{\partial \theta} g(\theta, \tilde b_{i \k}, \pi_i, \Psi) \right] \left[ \sum_{\k} \omega(\theta, \tilde b_{i \k}, \pi_i, \Psi) \frac{\partial}{\partial \theta} g(\theta, \tilde b_{i \k}, \pi_i, \Psi) \right]^\top
\end{aligned}
$$

When $N_{GQ} = 1$, the AGQ approximation is equivalent to a Laplacian approximation.
